
# Table of Contents

1.  [Annotation resources - `ensembldb`](#org2d40964)
    1.  [Annotation of genomic regions](#org148e20e)
    2.  [`AnnotationFilter`: basic classes for filtering annotation resources](#orgdc2698c)
    3.  [Filtering `EnsDb` databases](#org8249192)
    4.  [Additional `ensembldb` capabilities](#org2e50b34)
    5.  [Finally&#x2026;](#org5be21c5)



<a id="org2d40964"></a>

# Annotation resources - `ensembldb`

**CSAMA2018**

**Johannes Rainer** (Eurac Research, Italy)
johannes.rainer@eurac.edu
github/twitter: *jotsetung*


<a id="org148e20e"></a>

## Annotation of genomic regions

-   Annotations for genomic features (genes, transcripts, exons) provided by
    `TxDb` (`GenomicFeatures`) and `EnsDb` (`ensembldb`) databases.
-   `EnsDb`:
    -   Designed for Ensembl-based annotations.
    -   One database per species and Ensembl release.
-   Extract data using methods:
    -   `genes`
    -   `transcripts`
    -   `exons`
    -   `txBy`
    -   `exonsBy`
    -   &#x2026;
-   Results are returned as `GRanges` or `GRangesList` objects.
-   <span class="underline">Example</span>: get all gene annotations from an `EnsDb`:
    
      ```{r   }
      ## Load the database for human genes, Ensembl release 86.
      library(EnsDb.Hsapiens.v86)
      edb <- EnsDb.Hsapiens.v86
    
      ## Get all genes from the database.
      gns <- genes(edb)
    
      head(gns)
    
      ## Metadata columns; gene biotype
      head(gns$gene_biotype) 
      ```


<a id="orgdc2698c"></a>

## `AnnotationFilter`: basic classes for filtering annotation resources

-   Extracting the full data not always required: filter databases.
-   `AnnotationFilter` provides basic classes and concepts for filtering.
-   One filter class for each annotation type/database column.
-   <span class="underline">Example</span>: create filters
    
      ```{r   }
      ## Create filter using the constructor function
      gnf <- GenenameFilter("BCL2", condition = "!=")
      gnf
    
      ## Create using a filter expression
      gnf <- AnnotationFilter(~ genename != "BCL2")
      gnf
    
      ## Combine filters
      AnnotationFilter(~ seq_name == "X" & gene_biotype == "lincRNA") 
      ```


<a id="org8249192"></a>

## Filtering `EnsDb` databases

-   <span class="underline">Example</span>: what filters can we use?
    
      ```{r   }
      ## List all supported filters by an EnsDb
      supportedFilters(edb) 
      ```

-   Pass filter(s) to `EnsDb` methods with the `filter` parameter or use the
    `filter` function to subset the database.
-   <span class="underline">Example</span>: get all transcripts for the gene *BCL2*.
    
      ```{r   }
      ## Get all transcripts for BCL2
      txs <- transcripts(edb, filter = ~ genename == "BCL2")
      txs
    
      ## Combine filters: only protein coding tx for the gene
      txs <- transcripts(edb, filter = ~ genename == "BCL2" &
    			      tx_biotype == "protein_coding")
    
    
      ## For the pipe lovers:
      library(magrittr)
      txs <- edb %>%
          filter(~ genename == "BCL2" & tx_biotype == "protein_coding") %>%
          transcripts 
      ```

-   Filters speed up queries.


<a id="org2e50b34"></a>

## Additional `ensembldb` capabilities

-   `EnsDb` contain also protein annotation data:
    -   Protein sequence.
    -   Annotation to Uniprot accessions.
    -   Annotation of all protein domains within the protein sequences.
-   Functionality to map coordinates:
    -   `genomeToTranscript`
    -   `genomeToProtein`
    -   `transcriptToGenome`
    -   `transcriptToProtein`
    -   `proteinToGenome`
    -   `proteinToTranscript`


<a id="org5be21c5"></a>

## Finally&#x2026;

**Thank you for your attention!**

**&#x2026; and good luck to the brave runners &#x2026;**

