---
title: "Lab 1: Introduction to _R_ and _Bioconductor_"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteIndexEntry{Lab 1: Introduction to R and Bioconductor}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Original Authors: Martin Morgan, Sonali Arora<br />
Presenting Authors: [Martin Morgan][], [Lori Shepherd][]</br >
Date: 9 July, 2018
Back: [Monday labs](lab-1-intro-to-r-bioc.html)

[Martin Morgan]: mailto: Martin.Morgan@RoswellPark.org
[Lori Shepherd]: mailto: Lori.Shepherd@RoswellPark.org

This lab has four parts. Each part consists of informative material
followed by short or intermediate length exercises and case
studies. The material introduces or refreshes your knowledge of _R_,
and walks through some of the central classes and annotation resources
for working with high-throughput sequence data in _Bioconductor_.

It's recommended that you choose two or three of the following
sections, and spend about 45 minutes to an hour on each. Take time to
work through the code in the material as well as the exercises, asking
questions of your colleagues and the TAs as needed. If you're
relatively new to _R_, then consider doing sections 1 and 3. If you're
new to _Bioconductor_, then skim the second and focus on the third and
fourth sections.

Download these files:

- [ALLphenoData.tsv](./ALLphenoData.tsv)
- [BRFSS-subset.csv](./BRFSS-subset.csv)
- [symgo.csv](./symgo.csv)

Choose the material you'd like to cover:

1.1 [Introduction to _R_](L1.1-r-intro-morgan.html) -- Gain confidence
    working with base R commands and data structures.

1.2 [Introduction to _Bioconductor_](L1.2-bioc-intro-morgan.html) --
    An overview of software available in _Bioconductor_.

1.3 [Data Representations](L1.3-bioc-data-representation-morgan.html)
    -- Learn the essentials of _Bioconductor_ data structures

1.4 [Annotation](L1.4-bioc-annotation-morgan.html) -- Learn about
    _Bioconductor_ resources for gene and genome annotation.

