---
title: "Lab 1: Introduction to _R_ and _Bioconductor_"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteIndexEntry{Lab 1: Introduction to R and Bioconductor}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Original Authors: Martin Morgan, Sonali Arora<br />
Presenting Author: Martin Morgan (<a
  href="mailto:martin.morgan@roswellpark.org">martin.morgan@roswellpark.org</a>)<br />
Date: 11 July, 2016

This lab has four parts. Each part consists of informative material
followed by short or intermediate length exercises and case
studies. The material introduces or refreshes your knowledge of _R_,
and walks through some of the central classes and annotation resources
for working with high-throughput sequence data in _Bioconductor_.

It's recommended that you choose two of the following sections, and
spend about 45 minutes on each. Take time to work through the code in
the material as well as the exercises, asking questions of your
colleagues and the TAs as needed. If you're relatively new to _R_,
then consider doing sections 1 and 3. If you're new to _Bioconductor_,
then skim the second and focus on the third and fourth sections.

Download these files:

- [ALLphenoData.tsv](../extdata/ALLphenoData.tsv)
- [BRFSS-subset.csv](../extdata/BRFSS-subset.csv)
- [symgo.csv](../extdata/symgo.csv)

Choose the material you'd like to cover:

1.1 [Introduction to _R_](L1.1-r-intro-morgan.html) -- Gain confidence
    working with base R commands and data structures.

1.2 [Introduction to _Bioconductor_](L1.2-bioc-intro-morgan.html) --
    An overview of software available in _Bioconductor_.

1.3 [Data Representations](L1.3-bioc-data-representation-morgan.html)
    -- Learn the essentials of _Bioconductor_ data structures

1.4 [Annotation](L1.4-bioc-annotation-morgan.html) -- Learn about
    _Bioconductor_ resources for gene and genome annotation.

