---
title: "Evening Session: Using GPUs for R/Bioc"
author: Nitesh Turaga, Martin Morgan
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
    eval = as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache = as.logical(Sys.getenv("KNITR_CACHE", "TRUE"))
)
```

# What is a GPU?

- The graphics processing unit, or GPU, has become one of the most
  important types of computing technology to dramatically accelerate
  workloads in high performance computing (HPC), deep learning and
  more.

- Designed for parallel processing

- GPUs were originally designed to accelerate the rendering of 3D graphics

- While CPUs have continued to deliver performance increases through
  architectural innovations, faster clock speeds, and the addition of
  cores, GPUs are specifically designed to accelerate computer
  graphics workloads.
  
-  GPUs are more programmable than ever before, affording them the
   flexibility to accelerate a broad range of applications that go
   well beyond traditional graphics rendering.
   
## GPU for Machine learning

- Some of the most exciting applications for GPU technology involve AI
and machine learning.

- GPUs incorporate an extraordinary amount of computational
capability, they can deliver incredible acceleration in workloads that
take advantage of the highly parallel nature of GPUs, such as image
recognition. 

- Many of today’s deep learning technologies rely on GPUs working in
conjunction with CPUs.

## GPGPU - General Purpose GPU

Applications in Bioinformatics - https://en.wikipedia.org/wiki/General-purpose_computing_on_graphics_processing_units#Bioinformatics 


## Packages in Bioconductor using deep learning

`ttgsea` - Tokenizing Text of Gene Set Enrichment Analysis
`DeepPINCS` - Protein Interactions and Networks with Compounds based on Sequences using Deep Learning
`VAExprs` - Generating Samples of Gene Expression Data with Variational Autoencoders
`GenProSeq`	-  Generating Protein Sequences with Deep Generative Models

## GPUMagic - Jiefei Wang and Martin Morgan

The gpuMagic package is designed for ultilizing the computing power of
modern GPU. The package has the ability to compile R code and
implement the code on a GPU kernel. Therefore, it is easy for the user
to design their own algorithm in pure R language and run the code in
GPU. The package uses openCL as its parallel language so
parallelization using the CPU is also supported.

https://bioconductor.org/packages/devel/bioc/vignettes/gpuMagic/inst/doc/Quick_start_guide.html

## 
