---
title: "Lab 1.5: Lecture Activities"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteIndexEntry{Lab 1.5: Lecture Activities}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Lecture 1

- See parts 2-4 of the lab.

Lecture 2

1. R has the value `pi` defined in the base package. Type `pi` to see
   this value. Assign `pi = "apple"` in the global environment. Use
   `search()` to see how R looks for symbols. Type `pi` at the command
   prompt to get the value of `pi` you assigned. How do you get the
   value of `pi` defined in the base environment, without actually
   removing `pi` from the global environment (hint: `?"::"`)

2. (Advanced) The `lm()` function fits a linear model and returns an
   'S3' class `lm` (e.g., `fit <- stats::lm(...); class(fit)` returns
   "lm"). Most Bioconductor classes are 'S4', eg., `dna <-
   Biostrings::DNAStringSet(c("AAA", "CCC")); class(dna)`. Use
   `getClass("DNAStringSet")` to see the S4 class definition. What is
   the class definition of an 'lm' object?

Lecture 3

1. Simulate tossing a fair coin 100 times. Repeat for 1000 replicates
   to approximate the null distribution. Compare to a binomial distribution.
   
2. Compare the probability of observing 6 heads of 10 tosses, 60 of
   100, and 600 of 1000 tosses. What lesson do we learn about sample
   size and power?
   
3. Construct the figure Wolfgang showed, plotting the quantiles of
   `1 - p` against the t statistic. Validate, e.g., through
   simulation, that the p values are uniformly distributed. Simulate
   correlated data and demonstrate the consequences for `p`.
   
Lecture 4

1. Simulate a single experiment of, e.g., 'height' between two groups,
   where the mean differs. Perform a t-test using `t.test()` to
   demonstrate that the means differ. Fit a linear model using `lm()`,
   and test for differences between groups. Are the p-values from the
   t-test and linear model the same? Should they be? What needs to be
   done to make the tests the same?
   
2. Visit the 'anscombe' help page `?anscombe` and work through the
   example section.

3. Recreate the analysis of Mr. Whiteside's home insulation project.
