---
title: "PBMC Cluster9"
author: "RG"
date: "2023-06-13"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("isotree")
library("scater")
load("exprs9.rda")
```

## The plot of PC4 vs PC1 ...

- suggests that PC4 might be two groups
- if it consists of two sub-groups then if we compute distances between all pairs of points we should have some sort of multi-modal distribution
- here we are going to use the isolation forest approach

## Let's use isolation forests to see if we can better understand cluster 9

- first we will drop genes for which we have fewer than 15 samples with that gene observed
- this could be problematic if our variability
```{r clean, echo=TRUE}
rS = apply(exprs9, 1, function(x) sum(x==0))
exprs9a = exprs9[rS<100,]
iF = isolation.forest(t(exprs9a))
preds=predict(iF, t(exprs9a))


```

## And ...
 - we see a bimodal distribution suggesting that there are two types of points in this cluster
 - just as we had seen in our previous analysis
```{r plotiF, echo=FALSE}
plot(density(preds), main="Isolation Forest Predictions",xlab="iF Score")
```

## Do we have outliers?

- the basic advice is that an outlier has values close to 1
```{r outliers, echo=TRUE}
summary(preds)
```
- the values are all less than 1/2 - so no outliers seem to be present

## Other things we could do...

- build an isolation forest for all the data in the PBMC data set
- compute predictions for all of the cells
- look at boxplots of the predictions for each cluster (or maybe stacked density plots)
- are the proposed clusters homogeneous with respect to the isolation forest notion of distance
