---
title: "19.2: Gene Set Enrichment Analysis"
author:
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{19.2: Gene Set Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: true
---

```{r style, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE"))
)
options(width = 75)
```

# Theory

See [slides][].

[slides]: ./lecture-19-gene-set-enrichment.pdf

A recent [tweet][] provides a nice summary of efforts to benchmark
gene set enrichement analysis methods using the [GSEABenchmarkR][]
package.

# Practice

```{r, message = FALSE, echo = FALSE}
library(DESeq2)
library(airway)
library(dplyr)
library(org.Hs.eg.db)
library(GO.db)
library(limma)
```

Data input and massage

```{r}
library(airway)
data(airway)
airway$dex <- relevel(airway$dex, "untrt")
```

Differential expression analysis

```{r, message = FALSE}
library(DESeq2)
des <- DESeqDataSet(airway, design = ~ cell + dex)
des <- DESeq(des)
res <- results(des)
```

Transition to tidy data

```{r}
library(dplyr)
library(tibble)
tbl <- res %>%
    as.data.frame() %>%
    as_tibble(rownames = "ENSEMBL")
tbl
```

## Example: hypergeometric test using [limma][]`::goana()`

Requires ENTREZ identifiers

```{r, message=FALSE}
library(org.Hs.eg.db)
tbl <- tbl %>%
    mutate(
        ENTREZID = mapIds(org.Hs.eg.db, ENSEMBL, "ENTREZID", "ENSEMBL")
    ) %>%
    dplyr::select(ENSEMBL, ENTREZID, everything())
tbl
```

Universe -- must be testable for DE

```{r}
tbl <- tbl %>%
    filter(!is.na(padj), !is.na(ENTREZID))
tbl
```

[limma][]`::goana()` -- Hypergeometric

```{r}
library(limma)
go <-
    goana(tbl$ENTREZID[tbl$padj < .05], tbl$ENTREZID, "Hs") %>%
    as_tibble(rownames = "GOID") %>%
    dplyr::select(GOID, Ont, everything())
go
```

What GO id's are most differentially expressed? (Why do these gene
sets seem to have a large size, `N`?)

```{r}
go %>% arrange(P.DE)
```

Sanity check?

```{r}
go %>%
    filter(grepl("glucocorticoid", Term)) %>%
    arrange(P.DE)
```

What genes in set?

```{r, message=FALSE}
genesets <-
    AnnotationDbi::select(org.Hs.eg.db, tbl$ENTREZID, "GOALL", "ENTREZID") %>%
    as_tibble() %>%
    dplyr::select(GOID = GOALL, Ont = ONTOLOGYALL, ENTREZID) %>%
    distinct() %>%
    arrange(Ont, GOID)
genesets
```

# Provenance

```{r}
sessionInfo()
```

[limma]: https://bioconductor.org/packages/limma
[GSEABenchmarkR]: https://bioconductor.org/packages/GSEABenchmarkR
[tweet]: https://twitter.com/LeviWaldron1/status/1142092301403115521
