---
title: "Process SC Data"
author: "RG"
date: "2023-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Download and Process the Data

FIXME:  we should update this to more modern workflows.
Somehow there is no filtering going on, wrt, the counts matrix, which should
be made smaller.

I am not convinced that denoisePCA is doing what the author thinks it is.
For the PCA the genes provide direction, but the variation we are modeling is wrt between cell
distances (not about variation within gene across cells, which seems to be
what is being used).

```{r, echo=FALSE}
library(DropletTestFiles)
library(GSEABase)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)
```

## Some Ideas from Mardia Kent and Bibby

- one measure of variable importance can be ascertained from the PCs.  For PC1, for example, any feature (gene) that has a large *loading* on PC1 is likely to be important in understanding the variation in that direction

- one could look at the highly variable genes and see what their loadings are in each of the PCs


```{r, echo=FALSE}
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")

```

```{r, echo=FALSE}
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]

unfiltered <- sce.pbmc

stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]

summary(high.mito)
```

scDBLfinder

```{r, echo=FALSE}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- high.mito

gridExtra::grid.arrange(
    plotColData(unfiltered, y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, y="subsets_Mito_percent",
        colour_by="discard") + ggtitle("Mito percent"),
    ncol=2
)

```

```{r, echo=FALSE}
library("scran")
library("scater")
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
save(clusters, file="clusters.rda")
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

summary(sizeFactors(sce.pbmc))
```

##what we do - at CCB
- drop cells with too many Mito reads
- drop genes found in too few cells
- drop cells with too few genes detected
- 
basic PCA on the 2000 most variable genes. We’ll save the top 50 PCs for each sample.
UMAP on all the gene expression values


```{r, echo=FALSE}
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")
ncol(reducedDim(sce.pbmc, "PCA"))
```
Now let's have a look at what we got

## Break

```{r}
##Extract the computed PCs
pcs <- getDenoisedPCs(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)
##components is the matrix of rotated per cell 
dim(pcs$components)
##and rotation is the matrix of PC directions
dim(pcs$rotation)
```

And now save the appropriate things for loading up later.
```{r savestuff, echo=TRUE}
save(pcs, file="pcs.rda")
save(sce.pbmc, file="sce.pbmc.rda")


```
