---
title: "Speeding up R code with Renjin"
author: "Wolfgang Huber"
output: BiocStyle::html_document
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Load the renjin R package

Renjin in general, and the R package that contains renjin are described at http://docs.renjin.org/en/latest/package. It can be installed via the instructions given there (it is currently not on CRAN). Note that it requires a working instance of the R package `rJava` installed.

```{r renjin}
library("renjin")
```

# Computing pi via the Leibniz formula

We use this as an example of an algebra-heavy computation that involves a lengthy loop.
We approximate the below infinite sum by one with a large value of `m`.

\begin{equation}
\frac{\pi}{4} =  \lim_{m\to\infty}\sum_{n=0}^{m} \frac{(-1)^n}{2n+1}
\end{equation}

```{r oldJIT1, echo = FALSE}
oldJIT <- compiler::enableJIT(0)
```

```{r compute_pi}
compute_pi <- function(m) {
    s = 0
    sign = 1
    for (n in 0:m) {
        s = s + sign / (2 * n + 1)
        sign = -sign
    }
    4 * s
}
```

## R's built-in byte code compiler

```{r}
compute_pi_bcc <- compiler::cmpfun(compute_pi) 
```

## Vectorization 

Vectorization is often a very appropriate way to get rid of explicit loops in R, and thus to speed up code, and even make it more readable and more robust. However, vectorization also has two potential drawbacks:

- When starting from a loop-based version, it requires manual rewriting of code. In the example here, that is simple, even trivial, but for more complex calculations, e.g., if the computations in one loop iteration depend on those from previous iterations, this may not always be easy or possible. (An example for that would be an iterative optimization algorithm.)

- It requires the allocation of large vectors; in the example below, of the vector `n` as well as several temporary expressions that arise in the evaluation of the second line of the below function. A double precision float vector of length $5\times10^8$ requires 3.7 GB.

```{r vectorized}
compute_pi_vec <- function(m) {
  n <- 0:m
  4 * sum((-1)^n / (2 * n + 1))
}
```

## Rcpp

Iterative algorithms can often be implemented in C / C++. However, one has to be able to program in multiple languages and the development process can be slow.

```{r compute_pi_cpp}
library(Rcpp)
compute_pi_cpp <- cppFunction("
    double compute_pi_cpp(int m) {
        double s = 0, sign = 1;
        for (int n = 0; n <= m; ++n) {
            s = s + sign / (2 * n + 1);
            sign = -sign;
        }
        return 4 *s;
    }
")
```

# Benchmark

```{r benchmark}
pis <- rep(NA_real_, 5)
m <- 5e8
```

## Ordinary uncompiled R function

```{r b1}
system.time(pis[1] <- compute_pi(m))
```

<!-- Note that we **reduced `m` by a factor of 10** here to limit runtime. -->

## Compiled R function

```{r b2}
system.time(pis[2] <- compute_pi_bcc(m))
```

Indeed, R will already compile many functions by default, automatically, without your explicit calling of the compiler. This behavior depends on a global state setting of R that is accessed by the `enableJIT` function in the `compiler` package. For demonstration purposes, we have above disabled the automatic compilation.

## renjin 

```{r b3}
system.time(pis[3] <- renjin(compute_pi(m)))
```

## Vectorized

```{r b4}
system.time(pis[4] <- compute_pi_vec(m))
```

## Rcpp

```{r b5}
system.time(pis[5] <- compute_pi_cpp(m))
```

## Check that the results are the same

```{r check}
print(pis)
max(abs(pis[-length(pis)] - pis[length(pis)]))
```

<!-- .. at least for the 2nd to 4th calculation; for the 1st we had used a smaller `m`. -->

```{r oldJIT2, echo = FALSE}
compiler::enableJIT(oldJIT)
```

# Acknowledgements

Parts of the development of renjin was funded by the European Commission through the Horizon 2020 project SOUND: http://www.sound-biomed.eu

# Session Info

```{r session}
devtools::session_info()
```
