---
title: "Profiling demonstration for CSAMA 2023"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Profiling demonstration for CSAMA 2023}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Overview

Understanding the resource usage of a workflow is important
for managing compute and storage resources.

There are tools in R to profile memory consumption.  CPU,
disk and storage activities are also important to measure.
For linux systems, the [collectl programs](https://collectl.sourceforge.net/)
conveniently measure consumption in user-selectable time slices.

# Demonstration

## Data production and clustering by various algorithms

```{r demo1}
library(Rcollectl)
library(BiocParallel)
cl = cl_start()
cl_timestamp(cl,"make 5e6x100 data")
 mat = matrix(rnorm(5e6), ncol=100)
cl_timestamp(cl,"start 10k dist")
 dd = dist(mat[seq_len(1e4),])
cl_timestamp(cl,"start hclust")
 z = hclust(dd)
 library(bluster)
cl_timestamp(cl,"start 10k clusterRows")
 ee = clusterRows(mat[seq_len(1e4),], KNNGraphParam(BPPARAM=MulticoreParam()))
cl_timestamp(cl,"start 20k dist")
 dd = dist(mat[seq_len(2e4),])
cl_timestamp(cl,"start hclust")
 z = hclust(dd)
 library(bluster)
cl_timestamp(cl,"start 20k clusterRows")
 ee = clusterRows(mat[seq_len(2e4),], KNNGraphParam(BPPARAM=MulticoreParam()))
cl_timestamp(cl,"start 20k leiden")
 ee = clusterRows(mat[seq_len(2e4),], KNNGraphParam(BPPARAM=MulticoreParam(),cluster.fun="leiden"))
cl_stop(cl)
```

## Graphical report

```{r rep}
path <- cl_result_path(cl)
plot_usage(cl_parse(path)) +
   cl_timestamp_layer(path) +
   cl_timestamp_label(path) +
   ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust=1))
```

