---
title: "Evening Session: Using _R_ and _Bioconductor_ with _Docker_"
author: Nitesh Turaga
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
    eval = as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache = as.logical(Sys.getenv("KNITR_CACHE", "TRUE"))
)
```

# What is Docker?

<!-- A Unix operating system is broken into two primary components, the **kernel space**, and the **user space**. -->

<!-- The **Kernel** talks to the hardware, and provides core system features.  -->

<!-- The **user space** is the environment that most people are most familiar with. It is where applications, libraries and system services run. -->




# Why use Docker for Bioconductor?

- Easy install of packages 

# How to use Docker for Bioconductor?

# What are the current Bioconductor images available?

**Stable release** of Bioconductor - R 4.2.0 , Bioc 3.15.0

`bioconductor/bioconductor_docker:RELEASE_X_Y`

**Development version** - R 4.2.0, Bioc 3.16.0

`bioconductor/bioconductor_docker:devel`

# Using Bioconductor docker images

## Basics 

Run the docker image

```
docker run \
     -e PASSWORD=bioc \
     -p 8787:8787 \
     bioconductor/bioconductor_docker:devel
```

## Mount volumes 


## Run the docker image with docker-compose

```
docker-compose up
```

## Extend docker images

### Example docker image

```
docker pull nitesh1989/bioconductor_msmb:RELEASE_3_15
```

## Singularity

Singularity was created to run complex applications on HPC clusters in a simple, portable, and reproducible way.

Advantages: 


Doing the same thing as above using Singularity:

### Run

### Mount 


## Resources and links

https://bioconductor.org/help/docker/ 

https://hub.docker.com/r/bioconductor/bioconductor_docker

