---
title: "Lecture 1 -- Introduction to R"
author: "Martin Morgan <Martin.Morgan@RoswellPark.org>"
date: "12 June 2017"
output:
  BiocStyle::html_document2:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{Lecture 1 -- Introduction to R}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

# History

- Statistical programming language
- 'Free' software: no cost, open source, broad use
- Extensible: packages (>11,000 on [CRAN][], >1300 on [Bioconductor][])
- Key features
    - Intrinsic _statistical_ concepts
    - _Vectorized_ computation
    - 'Old-school' scripts rather than graphical user interface -- great
      for reproducibility!
    - (Advanced) _copy-on-change_ semanatics

# Vectors and data frames

```{r}
1 + 2
x = c(1, 2, 3)
1:3             # sequence of integers from 1 to 3
x + c(4, 5, 6)  # vectorized
x + 4           # recycling
```

Vectors

- `numeric()`, `character()`, `logical()`, `integer()`, `complex()`, ...
- `NA`: 'not available'
- `factor()`: values from restricted set of 'levels'.

Operations

- numeric: `==`, `<`, `<=`, `>`, `>=`, ...
- logical: `|` (or), `&` (and), `!` (not)
- subset: `[`, e.g., `x[c(2, 3)]`
- assignment: `[<-`, e.g., `x[c(1, 3)] = x[c(1, 3)]`
- other: `is.na()`

Functions

```{r}
x = rnorm(100)
y = x + rnorm(100)
plot(x, y)
```

- Many!

`data.frame`

```{r}
df <- data.frame(Independent = x, Dependent = y)
head(df)
df[1:5, 1:2]
df[1:5, ]
plot(Dependent ~ Independent, df)  # 'formula' interface
```

- List of equal-length vectors
- Vectors can be of different type
- Two-dimensional subset and assignment
- Column access: `df[, 1]`, `df[, "Indep"]`, `df[[1]]`,
  `df[["Indep"]]`, `df$Indep`
  
Exercise: plot only values with `Dependent > 0`, `Independent > 0`

1. Select rows

    ```{r}
    ridx <- (df$Dependent > 0) & (df$Independent > 0)
    ```

2. Plot subset

    ```{r}
    plot(Dependent ~ Independent, df[ridx, ])
    ```
    
3. Skin the cat another way

    ```{r}
    plot(
        Dependent ~ Independent, df,
        subset = (Dependent > 0) & (Independent > 0)
    )
    ```
    
# Analysis: functions, classes, methods

```{r}
fit <- lm(Dependent ~ Independent, df)  # linear model -- regression
anova(fit)                              # summary table
plot(Dependent ~ Independent, df)
abline(fit)
```

- `lm()`: plain-old function
- `fit`: an _object_ of class "lm"
- `anova()`: a _generic_ with a specific _method_ for class "lm"

```{r}
class(fit)
methods(class="lm")
```

# Help!

```{r, eval=FALSE}
?"plot"          # plain-old-function or generic
?"plot.formula"  # method
```

# Packages


```{r}
library(ggplot2)
ggplot(df, aes(x = Independent, y = Dependent)) +
    geom_point() + geom_smooth(method = "lm")
```

- General purpose: >11,000 packages on [CRAN][]
- Gain contributor's domain expertise _and_ weird (or other) idiosyncracies
- _Installation_ (once only per computer) versus _load_ (via
  `library(ggplot2)`, once per session)

[CRAN]: https://cran.r-project.org
[Bioconductor]: https://bioconductor.org


# End matter

## Session Info

```{r}
sessionInfo()
```

## Acknowledgements

Research reported in this tutorial was supported by the National Human
Genome Research Institute and the National Cancer Institute of the
National Institutes of Health under award numbers U41HG004059 and
U24CA180996.

This project has received funding from the European Research Council
(ERC) under the European Union's Horizon 2020 research and innovation
programme (grant agreement number 633974)
